A Theory of Bidirectional Type Synthesis for Simple Types

### 2023-07-24 tex/sec1-intro.tex
[ tex ] Remove defensive phrases; a comment on annotatability

Instead of focusing solely on a bidirectional type synthesiser like \citet{Wadler2022}~(\cref{sec:PLFA}), our theory centres around a general specification of type synthesis for an ordinary type system where judgements are not assigned modes, and fills in the gap between language formalisation frameworks based on such type systems~(\cref{sec:language-formalisation}) and raw terms produced by parsers.
Working similarly to those frameworks, our theory is parametrised by a \emph{bidirectional binding signature}, which extends \citeauthor*{Aczel1978}'s binding signature and describes a language with a bidirectional type system, analogously to a grammar describing a formal language in the theory of parsing.
As a first step, we confine our theory to \emph{syntax-directed} systems, meaning that for every raw term construct, the ordinary type system has exactly one typing rule, which is assigned modes in exactly one way in the bidirectional type system.%
\footnote{It is reasonable to assume syntax-directedness (especially as a first step) given that it is a standard idea to re-cast non-syntax-directed typing rules into their syntax-directed form to derive a type synthesiser~\citep{Peyton-Jones2007}.}
With this assumption, using a single bidirectional binding signature we can specify the raw terms, the ordinary type system, and the bidirectional type system of a language.
Then, if the signature is \emph{mode-correct}~\citep[Section~3.1]{Dunfield2021}, we can instantiate a type synthesiser that is formulated similarly to \varcitet{Wadler2022}{'s} with respect to the ordinary type system, and performs bidirectional type synthesis internally.
Our theory thus works as a verified type-synthesiser generator, analogous to a parser generator.

>>>>

Instead of focusing solely on a bidirectional type synthesiser like \citet{Wadler2022}~(\cref{sec:PLFA}), our theory centres around a general specification of type synthesis for an ordinary type system where judgements are not assigned modes, and fills in the gap between language formalisation frameworks based on such type systems~(\cref{sec:language-formalisation}) and raw terms produced by parsers.
Working similarly to those frameworks, our theory is parametrised by a \emph{bidirectional binding signature}, which extends \citeauthor*{Aczel1978}'s binding signature and describes a language with a bidirectional type system, analogously to a grammar describing a formal language in the theory of parsing.
As a first step, we confine our theory to \emph{syntax-directed} systems, meaning that for every raw term construct, the ordinary type system has exactly one typing rule, which is assigned modes in exactly one way in the bidirectional type system.%
\footnote{It is reasonable to assume syntax-directedness given that it is a standard idea to re-cast non-syntax-directed typing rules into their syntax-directed form to derive a type synthesiser~\citep{Peyton-Jones2007}.}
With this assumption, using a single bidirectional binding signature we can specify the raw terms, the ordinary type system, and the bidirectional type system of a language.
Then, if the signature is \emph{mode-correct}~\citep[Section~3.1]{Dunfield2021}, we can instantiate a type synthesiser that is formulated similarly to \varcitet{Wadler2022}{'s} with respect to the ordinary type system, and performs bidirectional type synthesis internally.
Our theory thus works as a verified type-synthesiser generator, analogous to a parser generator.

### 2023-07-24 tex/sec4-pre-synthesis.tex
[ tex ] Remove defensive phrases; a comment on annotatability

On the other hand, when using a bidirectional type synthesiser to implement type synthesis~(\cref{thm:implementation}), our completeness is much simpler to use than annotatability, which requires the bidirectional type synthesiser to produce more complex evidence when the synthesis fails.
Annotatability also does not help the user to deal with missing annotations like (generalised) mode decoration does: although annotatability seems capable of determining where annotations are missing and even filling them in correctly, its antecedent requires a typing derivation, which is what the user is trying to construct and does not have yet.
Therefore we believe that the notion of annotatability is too complex for what it achieves, and our theory offers simpler and more useful alternatives.

>>>>

On the other hand, when using a bidirectional type synthesiser to implement type synthesis~(\cref{thm:implementation}), our completeness is much simpler to use than annotatability, which requires the bidirectional type synthesiser to produce more complex evidence when the synthesis fails.
Annotatability also does not help the user to deal with missing annotations like (generalised) mode decoration does: although annotatability seems capable of determining where annotations are missing and even filling them in correctly, its antecedent requires a typing derivation, which is what the user is trying to construct and does not have yet.
Therefore we believe that the notion of annotatability is too complex\LT{not clear what it means} for what it achieves, and our theory offers simpler and more useful alternatives.

### 2023-07-24 tex/sec5-synthesis.tex
[ tex ] Remove defensive phrases; a comment on annotatability

\begin{theorem}[Decidability of Bidirectional Type Synthesis and Checking] \label{thm:bidirectional-type-synthesis-checking}
  In a mode-correct bidirectional type system $(\Sigma, \Omega)$,
  \begin{enumerate}
    \item if\/ $\erase{\Gamma} |-_{\Sigma, \Omega} \isTerm{t}^{\syn}$, then it is decidable whether there is~$A$ such that $\Gamma |-_{\Sigma, \Omega} \isTerm{t} \syn A$;
    \item if\/ $\erase{\Gamma} |-_{\Sigma, \Omega} \isTerm{t}^{\chk}$, then it is decidable for any~$A$ whether $\Gamma |-_{\Sigma, \Omega} \isTerm{t} \chk A$.
  \end{enumerate}
\end{theorem}

>>>>

\begin{theorem}[Decidability of Bidirectional Type Synthesis and Checking] \label{thm:bidirectional-type-synthesis-checking}
  In a mode-correct bidirectional type system $(\Sigma, \Omega)$,
  \begin{enumerate}
    \item if\/ $\erase{\Gamma} |-_{\Sigma, \Omega} \isTerm{t}^{\syn}$, then it is decidable whether $\Gamma |-_{\Sigma, \Omega} \isTerm{t} \syn A$ for some $A$;
    \item if\/ $\erase{\Gamma} |-_{\Sigma, \Omega} \isTerm{t}^{\chk}$, then it is decidable for any~$A$ whether $\Gamma |-_{\Sigma, \Omega} \isTerm{t} \chk A$.
  \end{enumerate}
\end{theorem}
